Historical Soil Organic Carbon Budget

### 2020-04-02 ResultNotebooks/results_socbudget_old.Rmd
Update pdf and results

setConfig(cachefolder = "/home/kristine/WORK/R libraries/rd3mod_inputdata/cache/rev4.18.cellpanda/", forcecache = TRUE)
#setConfig(sourcefolder = "/home/kristine/mnt/rd3mod/inputdata/sources/")
setConfig(sourcefolder = "/home/kristine/WORK/R libraries/rd3mod_inputdata/sources/")

@@

```{r table dens climatezones}
    cell_GSOC      <- readSource("GSOC",  convert="onlycorrect")
    cell_WISE      <- readSource("WISE",  convert="onlycorrect")
    cell_SoilGrids <- readSource("SoilGrids",  convert="onlycorrect")

>>>>

setConfig(cachefolder = "/home/kristine/WORK/R libraries/rd3mod_inputdata/cache/rev4.18.cellpanda/", forcecache = TRUE)
#setConfig(cachefolder = "/home/kristine/mnt/rd3mod/inputdata/cache/rev4.43/", forcecache = TRUE)
setConfig(sourcefolder = "/home/kristine/mnt/rd3mod/inputdata/sources/")
#setConfig(sourcefolder = "/home/kristine/WORK/R libraries/rd3mod_inputdata/sources/")

@@

```{r table dens climatezones}
cell_GSOC      <- readSource("GSOC",  convert="onlycorrect")
cell_WISE      <- readSource("WISE",  convert="onlycorrect")
cell_SoilGrids <- readSource("SoilGrids", subtype="cstock_0_30", convert="onlycorrect")

### 2020-04-02 SOCbudgetPaper/SOCbudget.tex
Update pdf and results

We calculate annual land use type specific soil organic carbon stocks
for cropland, pastures and natural vegetation on half-degree resolution
for the period of 1965 to 2010 based on the following three steps: (1)
Calculating the land use (sub-)type specific steady-states and decay
rates for SOC stocks given the current biophysical, climatic and
agronomic conditions, (2) accounting for land conversation effects by
transferring SOC between land use types and (3) updating SOC stocks
based on the previous stock, the steady-state and the decay rate.

@@

We account for different carbon inputs sources depending on the land use
type (see table @ref(tab:datasourceinputs)). Following the IPCC
methodology carbon inputs are disaggregated into different structural
components depending on their lignin and nitrogen content (see
@ref(ipcc\_2019\_2019)). For each structural components the sum over all
carbon inputs sources is allocated to the respective SOC sub-pools. This
implies that not only the amount of carbon, but also their structural
composition is determining the effective inflow. Data sources for all
considered carbon inputs as well as for lignin and nitrogen content can
be found in table @ref(tab:datasourceinputs).

@@

Decay rates are influenced by climatic conditions, biophysical and
biochemical soil properties as well as management factors that vary over
time (t) and space (i). Following the steady-state method of the
refinement of the IPCC guidelines vol.~4 (\citet{ipcc_2019_2019}) for
mineral soils we consider temperature (temp), water (wat), sand fraction
(sf) and tillage (till) effects to spatially disaggregate default global
decay rates. Since the three different SOC sub-pools represent different
SOC characteristics, the global decay rates and the associated drivers
differ. Thus \(k_{sub}\) is given by

@@

For cropland we performed an assessment of tillage types and irrigation
conditions, whereas on pastures and natural vegetation, we assume
rainfed and non-tilled conditions. Data sources as well as considered
effects for each land use types are shown in table
@ref(tab:datasourcedecay). To account for variations of decay rates
within grid cells based on different tillage and irrigation regimes,
average rates based on area shares are calculated.

@@

We calculate SOC stocks based on the area shares of land use types (lut)
within the half-degree grid cells (i). If land is converted from one
land use type into others (!lut), a respective share of the SOC stocks
have to be reallocated as well. We account for land conversion at the
beginning of each time step \(t\) by calculating a preliminary stock
\(SOC_{lut,t*}\) via

@@

Carbon stocks \(SOC\) for each sub-pool (sub) converge towards the
calculate steady-state stock \(SOC^{eq}\) for each land-use types (lut),
each sub-pool (sub) and each annual time step (t) as represented in
equation @ref(eq:steadystate).

@@

@@

Following the tier 1 approach of the IPCC guidelines vol.~4
(\citet{ipcc_2006_2006}), stocks are estimated via stock change factors
given by the IPCC for the topsoil (0-30 cm) and based on measurement
data and expert choice. The factors differentiate between different crop
and management systems reflecting different dynamics under changed in-
and outflows without explicitly tracking these. This method is used
additionally to the Tier-2 approach to evaluate model accuracy.

@@

Land use patterns are based on the Land-Use Harmonization 2 (LUH2,
\citep{LUH2}) data set, which we aggregate from quarter degree to half
degree resolution. We disaggregate the five different cropland
subcategories of LUH2 into our 17 crop groups, assuming relative shares
for each gridcell based on the country and year specific area shares of
FAOSTAT data (\citep{FAOSTAT}). Land use transitions are calculated as
net area differences of the land use data on half-degree.

@@

Using half-degree yield data from LPJmL (\citep{LPJmL4_1}) as well as
half-degree cropland patterns (see @ref(\#sec:landuse)) we compile crop
functional type specific half-degree production patterns. We calibrate
cellular yields with one country-level calibration factor for each
crop-group to meet historical FAOSTAT production (\citep{FAOSTAT}).

Crop residues production is based on a revised methodology as described
in (\citep{bodirsky2012}) and will be explained in key aspects again due
to its central role for soil carbon modelling. Starting from crop
production estimates of the harvested organs and their respective crop
area, we estimate above-ground (ag) and below-ground (bg) residual
biomass using yield-dependent harvest indices and shoot:root ratios. We
assume that all bg residues are recycled to the soil, whereas ag
residues can be burned or harvested for other purposes such as feeding
animals (\citep{weindl}), fuel or for material use.

A fixed share of the ag residues is assumed to be burned on field, which
depends on the per-capita income of the country. Following
\citep{smil1999}) we assume 25\% burn share for low-income countries
according to worldbank definitions (\(<\,1000\,\tfrac{USD}{yr}\)), 15\%
for high-income (\(>\,10000\,\tfrac{USD}{yr}\) and linearly interpolated
shares for all middle-income countries.

Residue demand for feed is based on regional livestock specific feed
basekts (see \citep{weindl}) taking available ag residual biomass as
well as livestock production systems and thus again development state
into account. Here we aggregate crop groups to residue groups based on
their suitability as feed (see @ref(append:Table\_kcr2kres)).

We estimated a material use share for the straw residues group of 5\%
and a fuel demand share demand of 10\% in low income countries.

The remaining ag residues are assumend to be recycled to the soil.

@@

To transform dry matter estimates into carbon, we compiled crop and
residue group specific carbon to dry matter (c:dm) ratios (see appendix
table X) {[}citations{]} for harvested organs and applied a universal
c:dm ratio of 0.44 {[}LPJml???{]} for all above and below ground residue
biomass as well as for grass biomass.

@@

To disaggregate country level FAOSTAT livestock production values to
half-degree pattern, we use the following rule based assumptions which
was inspired by the approach of \citep{gilbert}.

For poultry, egg and monogastric meat production we

\begin{itemize}
\item
  divide into intensive and extensiv eproduction systems based on a
  per-capita (explain rules in more detail)
\item
  Based on the development state (thats not a fixed term) of a country,
  we devide poultry, egg and monogastric meat production into intensive
  and extensive systems.
\item
  Extensive production is allocated to built-up area share within a
  country based on the assumption that these animals are held in
  households, subsistence or small-holder farming systems.
\item
  Intensive production is distributed within a country using the crop
  production (excluding 2nd generation bioenergy crops) share, assuming
  that feed availability is the most driving factor for livestock
  location.
\end{itemize}

For dairy and ruminated meat production we split into pasture fed and
non-pasture fed fractions based on the aggregated feed mix within a
country (explain rules)

>>>>

Following the tier 2 approach of the refinement of IPCC guidelines
vol.~4 (\citet{ipcc_2019_2019}), we calculate annual land use type
specific soil organic carbon stocks for cropland, pastures and natural
vegetation on half-degree resolution for the period of 1965 to 2010
based on the following three steps: (1) Calculating the land use
(sub-)type specific steady-states and decay rates for SOC stocks given
the current biophysical, climatic and agronomic conditions, (2)
accounting for land conversation effects by transferring SOC between
land use types and (3) updating SOC stocks based on the previous stock,
the steady-state and the decay rate.

@@

We account for different carbon input sources depending on the land use
type (see table @ref(tab:datasourceinputs)). Following the IPCC
methodology carbon inputs are disaggregated into metabolic and
structural components depending on their lignin and nitrogen content
(see @ref(ipcc\_2019\_2019)). For each component the sum over all carbon
input sources is allocated to the respective SOC sub-pools via transfer
coefficients. This implies that not only the amount of carbon, but also
their structural composition is determining the effective inflow. Data
sources for all considered carbon inputs as well as for lignin and
nitrogen content can be found in table @ref(tab:datasourceinputs).

@@

The sub-pool specific decay rates are influenced by climatic conditions,
biophysical and biochemical soil properties as well as management
factors that vary over time (t) and space (i). Following the
steady-state method of the refinement of the IPCC guidelines vol.~4
(\citet{ipcc_2019_2019}) for mineral soils we consider temperature
(temp), water (wat), sand fraction (sf) and tillage (till) effects to
account for spatial variation of decay rates. Thus \(k_{sub}\) is given
by

@@

For cropland we distinguish the effect of different tillage (see
@ref(\#sec:tillage)) and irrigation (see @ref(\#sec:irrigation))
practices on decay rates, whereas on pastures and natural vegetation, we
assume rainfed and non-tilled conditions. Data sources as well as
considered effects for each land use types are shown in table
@ref(tab:datasourcedecay). To account for variations of decay rates
within each grid cell due to different tillage and irrigation regimes,
average rates based on area shares are calculated.

@@

We calculate SOC stocks based on the area shares of land use types (lut)
within the half-degree grid cells (i). If land is converted from one
land use type into others (!lut), the respective share of the SOC stocks
is reallocated. We account for land conversion at the beginning of each
time step \(t\) by calculating a preliminary stock \(SOC_{lut,t*}\) via

@@

Carbon stocks \(SOC\) for each sub-pool (sub) converge towards the
calculated steady-state stock \(SOC^{eq}\) for each land-use types
(lut), each sub-pool (sub) and each annual time step (t) as represented
in equation @ref(eq:steadystate).

@@

\subsubsection{Initialisation of SOC pools}

To initialize all SOC sub-pools we assume that

@@

Additionally to the tier 2 approach of the refinement of IPCC guidelines
vol.~4 (\citet{ipcc_2019_2019}), we also estimate SOC pools using the
IPCC tier 1 approach of IPCC guidelines vol.~4 (\citet{ipcc_2006_2006})
for comparison. Here, stocks are estimated via stock change factors
given by the IPCC for the topsoil (0-30 cm) and based on a review of
measurement data. The factors differentiate different crop and
management systems reflecting different dynamics under changed in- and
outflows without explicitly tracking these. The SOC stocks as thus
calulated

\begin{equation}
SOC_{\text{target}} = \sum_{c,s,i} SOC_{\text{ref}_{c,s,i}} \cdot F_{\text{LU}_{c,s,i}} \cdot F_{\text{MG}_{c,s,i}} \cdot F_{\text{I}_{c,s,i}} \cdot A_{c,s,i}
\label{eq:tier1}
\end{equation}

\textless!- also include an equation here --\textgreater{} \textless!-
even if there are just ``copied'' out of te guidelines so to say?
--\textgreater{} \textless!- more details will follow - how deep to go?
--\textgreater{}

@@

Land use patterns are based on the Land-Use Harmonization 2 (LUH2,
\citep{LUH2}) data set, which we aggregate from quarter degree to half
degree resolution. We disaggregate the five different cropland
subcategories (c3ann, c3per, c4ann, c4per, c3nfx) of LUH2 into our 17
crop groups, assuming relative shares for each gridcell based on the
country and year specific area shares of FAOSTAT data (\citep{FAOSTAT})
(see @ref(append:Table\_luh2fao2mag) for more details on the crop type
mapping). Land use transitions are calculated as net area differences of
the land use data on half-degree.

@@

Using half-degree yield data from LPJmL (\citep{LPJmL4_1}) as well as
half-degree cropland patterns (see @ref(\#sec:landuse)) we compile crop
group specific half-degree production patterns. We calibrate cellular
yields with one country-level calibration factor for each crop group to
meet historical FAOSTAT production (\citep{FAOSTAT}). Note that by using
physical cropland areas we account for multiple crop harvest events as
well as for fallows.

Crop residue production and management is based on a revised methodology
of (\citep{bodirsky2012}) and will be explained in key aspects again due
to its central role for soil carbon modelling. Starting from crop
production estimates of the harvested organs and their respective crop
area, we estimate above-ground (ag) and below-ground (bg) residual
biomass using yield-dependent harvest indices and shoot:root ratios. We
assume that all bg residues are recycled to the soil, whereas ag
residues can be burned or harvested for other purposes such as feeding
animals (\citep{weindl}), fuel or for material use.

A fixed share of the ag residues is assumed to be burned on field, which
depends on the per-capita income of the country. Following
\citep{smil1999}) we assume 25\% burn share for low-income countries
according to worldbank definitions (\(<\,1000\,\tfrac{USD}{yr}\)), 15\%
for high-income (\(>\,10000\,\tfrac{USD}{yr}\) and linearly interpolate
shares for all middle-income countries depending on their per-capita
income.

Residue demand for feed is based on country and livestock specific feed
basekts (see \citep{weindl}) taking available ag residual biomass as
well as livestock productivity into account. The three residue groups
(straw, high-lignin and low-lignin residues) of the feed baskets are
disaggregated to the crop groups, using\ldots. (see
@ref(append:Table\_kcr2kres)). \textless-- its a disggregation, no
--\textgreater{}

We estimated a material use share for the straw residues group of 5\%
and a fuel demand share demand of 10\% \textless-- only for straw, or
for all groups? --\textgreater{} in low income countries.

The remaining ag residues as well as all bg residues are assumend to be
recycled to the soil.

@@

To transform dry matter estimates into carbon, we compiled crop group
and plant part specific carbon to dry matter (c:dm) ratios (see
@ref(append:Table\_c2dm)) (\citet{Ma2018}).

@@

To disaggregate country level FAOSTAT livestock production values to
half-degree pattern, we use the following rule based assumptions which
were inspired by the approach of \citep{gilbert}.

For poultry, egg and monogastric meat production we divide into
intensive and extensive production systems based on a per-capita income
of the country. For low-income countries according to worldbank
definitions (\textless1000 USD/yr), we assume extensive production
systems. We located them according to (built-up areas shares\textbar{}
population shares) based on the idea that these animals are held in
households, subsistence or small-holder farming systems.

\begin{itemize}
\item
  Intensive production is distributed within a country using the crop
  production (excluding 2nd generation bioenergy crops) share, assuming
  that feed availability is the most driving factor for livestock
  location.
\item
  For dairy and ruminated meat production we
\item
  splot into pasture fed and non-pasture fed fractions based on the
  aggregated feed mix within a country (explain rules)
\end{itemize}

### 2020-04-02 SOCbudgetPaper/methods.Rmd
Update pdf and results

\begin{equation}
SOC_{t} = SOC_{t-1} - \frac{SOC_{t-1}-SOC_{\text{target}}}{D}
\intertext{with}
SOC_{\text{target}} = \sum_{c,s,i} SOC_{\text{ref}_{c,s,i}} \cdot F_{\text{LU}_{c,s,i}} \cdot F_{\text{MG}_{c,s,i}} \cdot F_{\text{I}_{c,s,i}} \cdot A_{c,s,i}}
\label{eq:tier1}
\end{equation}

@@

For dairy and ruminated meat production we

>>>>

\begin{equation}
SOC_{\text{target}} = \sum_{c,s,i} SOC_{\text{ref}_{c,s,i}} \cdot F_{\text{LU}_{c,s,i}} \cdot F_{\text{MG}_{c,s,i}} \cdot F_{\text{I}_{c,s,i}} \cdot A_{c,s,i}
\label{eq:tier1}
\end{equation}

@@

* For dairy and ruminated meat production we
